---
title: "Make test and train sets"
author: "Kaspar Bresser"
date: "04/10/2023"
output: 
  github_document:
    toc: true
  html_document: 
    theme: simplex
    highlight: pygments
    code_folding: show
    self_contained: TRUE
    toc: yes
    toc_float:
      collapsed: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message=FALSE,
                      autodep  = TRUE,
                      cache = FALSE,
                      fig.width = 5,
                      fig.asp = 0.618,
                      fig.align = "center")
```

Used the code below to generate the train and test sets used for the random forests.

```{r loading}
library(tidyverse)
library(furrr)
```

# Import MS detected peptides

Read the MS detected peptides into a nested tibble, then unnest and select the columns we want to continue with. Also, only keep unique cases. 

```{r read_ligands}
file.names <- list.files(("Data/MS_peptides")) 

HLA.ligands <- tibble(
  
  ligand.table = map( paste0("Data/MS_peptides/", file.names), read_tsv ),
  tumor = map_chr(str_split(file.names, "_"), 1)

  )

#HLA.ligands <- mutate(HLA.ligands, case_when(tumor))

HLA.ligands

HLA.ligands <- mutate(HLA.ligands, tumor = c("M026X1", "SKMEL95", "NKIRTIL006"))

# Define Columns to keep from the ligand tables
keep <- c("rna","ribo",  "swissprot_id", "ligand", "peptide" )

HLA.ligands %>% 
  unnest( cols = ligand.table) %>% 
  select(one_of(keep, "tumor")) %>% 
  distinct() -> HLA.ligands

HLA.ligands
```


Split test, train. 80/20

```{r}
HLA.ligands %>% 
  group_by(tumor) %>% 
  sample_frac(.8) -> train.ligands

HLA.ligands %>% 
  anti_join(train.ligands) -> test.ligands
```


import expression info

```{r}
file.names <- list.files("Data/protein_data/") 

expression.data <- tibble(
  
  expression.info = map( paste0("Data/protein_data/", file.names), read_tsv ),
  tumor = map_chr(str_split(file.names, "_"), 1)

  )

(expression.data <- unnest(expression.data, cols = expression.info) %>% select(!length))

```

import swissprotIDs that are in the library, import swissprot sequences, and combine with expression data

```{r}
"../2-random_forest_analyses/Data/Protein_per_Uniprot_entry_library_v2_RBP_GC_length_codon_AA_m6A_m5C_AtoI_m1A_m7G_CD8miRDB_PTM.csv" %>% 
  read_tsv() %>% 
  pull(Entry) -> IDs.in.library

uniprot <- read_csv("UniProt_reviewed_input.tsv") %>% rename(swissprot_id = "sequence_id")

uniprot %>% 
  filter(swissprot_id %in% IDs.in.library)  -> uniprot

expression.data %>% 
  nest(expr = !tumor) %>% 
  mutate(expr = map(expr, ~right_join(., uniprot))) %>% 
  unnest(expr) %>% 
  replace_na(list(rna = 0, ribo = 0)) -> expression.data

```


sample peptides for train data. 

```{r}
get_peptide <- function(tum, pep.len){
  train.ligands %>% 
    filter(tumor == tum) %>% 
    filter(nchar(peptide) == pep.len) %>% 
    pull(peptide) %>% 
    toupper() -> s
  
  num <- nrow(filter(train.ligands, tumor == tum & nchar(peptide) == pep.len))*10
  
  expression.data %>% 
    filter(tumor == tum) %>% 
    mutate(peptide = sequence) %>% 
    mutate(len = nchar(peptide)) %>% 
    slice_sample(n = num*2, weight_by = len, replace = T) %>% 
    rowwise() %>% 
    mutate(num = sample(1:(nchar(peptide)-pep.len), 1),
         peptide = str_sub(peptide, num, num+(pep.len-1))) %>% 
    ungroup() %>% 
    filter(!(peptide %in% s)) %>% 
    select(peptide, swissprot_id, tumor, rna, ribo) %>% 
    mutate(ligand = FALSE) %>% 
    filter(swissprot_id %in% IDs.in.library) %>% 
    distinct(peptide, .keep_all = T) %>% 
    slice_sample(n = num)
}



decoys <- map2_dfr(rep(c("M026X1", "SKMEL95", "NKIRTIL006"), 3), rep(c(9,10,11), each = 3), get_peptide)

train.set <- bind_rows(decoys, train.ligands)

train.set %>% 
  count(tumor, ligand, nchar(peptide))
```


sample peptides for test data. 

```{r}
get_peptide <- function(tum, pep.len){
  test.ligands %>% 
    filter(tumor == tum) %>% 
    filter(nchar(peptide) == pep.len) %>% 
    pull(peptide) %>% 
    toupper() -> s
  
  num <- nrow(filter(test.ligands, tumor == tum & nchar(peptide) == pep.len))*1000
  
  expression.data %>% 
    filter(tumor == tum) %>% 
    mutate(peptide = sequence) %>% 
    mutate(len = nchar(peptide)) %>% 
    slice_sample(n = num*4, weight_by = len, replace = T) %>% 
    rowwise() %>% 
    mutate(num = sample(1:(nchar(peptide)-pep.len), 1),
         peptide = str_sub(peptide, num, num+(pep.len-1))) %>% 
    ungroup() %>% 
    filter(!(peptide %in% s)) %>% 
    select(peptide, swissprot_id, tumor, rna, ribo) %>% 
    mutate(ligand = FALSE) %>% 
    filter(swissprot_id %in% IDs.in.library) %>% 
    distinct(peptide, .keep_all = T) %>% 
    slice_sample(n = num)
}


decoys <- map2_dfr(rep(c("M026X1", "SKMEL95", "NKIRTIL006"), 3), rep(c(9,10,11), each = 3), get_peptide)

test.set <- bind_rows(decoys, test.ligands)

test.set %>% 
  count(tumor, ligand, nchar(peptide))
```


Write out the table and peptide files


```{r}
write_tsv(train.set, "./Output/test_train_sets/HLA_train_set.tsv")

train.set %>% 
  transmute(pep.len = paste0(nchar(peptide), "AA"), peptide = peptide) %>% 
  nest(data = peptide) %>% 
  mutate(peptides = map(data, pull, peptide),
         peptides = set_names(peptides, pep.len)) %>% 
  pull(peptides) %>% 
  map2(names(.), ~write_lines(.x, paste0("./Output/test_train_sets/mel_Train_Peptides_", .y, ".tsv")))

write_tsv(test.set, "./Output/test_train_sets/HLA_test_set.tsv")

test.set %>% 
  transmute(pep.len = paste0(nchar(peptide), "AA"), peptide = peptide) %>% 
  nest(data = peptide) %>% 
  mutate(peptides = map(data, pull, peptide),
         peptides = set_names(peptides, pep.len)) %>% 
  pull(peptides) %>% 
  map2(names(.), ~write_lines(.x, paste0("./Output/test_train_sets/mel_Test_Peptides_", .y, ".tsv")))
```



