---
title: "run netMHCpan4.1"
author: "Kaspar Bresser"
date: "04/02/2024"
output: 
  github_document:
    toc: true
  html_document: 
    theme: simplex
    highlight: pygments
    code_folding: show
    self_contained: TRUE
    toc: yes
    toc_float:
      collapsed: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      eval = FALSE,
                      message=FALSE,
                      autodep  = TRUE,
                      cache = FALSE,
                      fig.width = 5,
                      fig.asp = 0.618,
                      fig.align = "center")

```

Used the code below to run netMHCpan on the train and test set peptides.

```{r loading}
library(tidyverse)
library(furrr)
```



## Run netMHCpan train set

Get the file names

```{r}
files <- paste0("Output/test_train_sets/", list.files("Output/test_train_sets/", pattern = "Train_Pep"))

files
```

check the peptide lengths for each file

```{r}
str_extract(files, "\\d+")
```



```{r}
HLA.alleles <-  c("HLA-A0201","HLA-A0101", "HLA-A2402", 
                  "HLA-B0801", "HLA-B1801","HLA-B5101", "HLA-B1501", "HLA-B4402", 
                  "HLA-C0501","HLA-C0701", "HLA-C1601", "HLA-C0304")
```


Run system command for netMHCpan, for each allele. Define function to get the arguments/flags for netMHCpan


```{r}
args_netMHC <- function(allele, file.name, peptide.len){
  c(
    paste0("-a ", sub("(.{7})(.*)", "\\1:\\2", allele)),
    paste0("-f ", file.name),
    "-p",
    "-rth 0.0",
    "-rlt 0.0",
    paste0("-l ", peptide.len),
    "-t -100"
  )
}

```

wrapper function to predict for each length

```{r}
netMHC_wrapper <- function(all.alleles, file.name){
  
  future_map(all.alleles, ~system2(command = "netMHCpan", 
                      args = args_netMHC(., file.name, str_extract(file.name, "\\d+")), 
                      stdout  = paste0("./Output/netMHCpan_predictions_train/netMHCpan_prediction_", .,
                                      "_", str_extract(file.name, "\\d+") , ".txt")))
  
}
```

Run netMHCpan, parallel for the alleles 

```{r}
plan(multisession, workers = 6)

map(files, ~netMHC_wrapper(HLA.alleles, .))
```


## Run netMHCpan test set

Get the file names

```{r}
files <- paste0("Output/test_train_sets/", list.files("Output/test_train_sets/", pattern = "Test_Pep"))

files
```

check the peptide lengths for each file

```{r}
str_extract(files, "\\d+")
```



```{r}
HLA.alleles <-  c("HLA-A0201","HLA-A0101", "HLA-A2402", 
                  "HLA-B0801", "HLA-B1801","HLA-B5101", "HLA-B1501", "HLA-B4402", 
                  "HLA-C0501","HLA-C0701", "HLA-C1601", "HLA-C0304")
```


Run system command for netMHCpan, for each allele. Define function to get the arguments/flags for netMHCpan


```{r}
args_netMHC <- function(allele, file.name, peptide.len){
  c(
    paste0("-a ", sub("(.{7})(.*)", "\\1:\\2", allele)),
    paste0("-f ", file.name),
    "-p",
    "-rth 0.0",
    "-rlt 0.0",
    paste0("-l ", peptide.len),
    "-t -100"
  )
}

```

wrapper function to predict for each length

```{r}
netMHC_wrapper <- function(all.alleles, file.name){
  
  future_map(all.alleles, ~system2(command = "netMHCpan", 
                      args = args_netMHC(., file.name, str_extract(file.name, "\\d+")), 
                      stdout  = paste0("./Output/netMHCpan_predictions_test/netMHCpan_prediction_", .,
                                      "_", str_extract(file.name, "\\d+") , ".txt")))
  
}
```

Run netMHCpan, parallel for the alleles 

```{r}
plan(multisession, workers = 6)

map(files, ~netMHC_wrapper(HLA.alleles, .))
```




