---
title: "combine netMHCpan result files"
author: "Kaspar Bresser"
date: "04/02/2024"
output: 
  github_document:
    toc: true
  html_document: 
    theme: simplex
    highlight: pygments
    code_folding: show
    self_contained: TRUE
    toc: yes
    toc_float:
      collapsed: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      eval = FALSE,
                      message=FALSE,
                      autodep  = TRUE,
                      cache = FALSE,
                      fig.width = 5,
                      fig.asp = 0.618,
                      fig.align = "center")

```

Used the code below to generate the train and test sets used for the random forests.

```{r loading}
library(tidyverse)
```

# Test data

## Clean up files

Define function to read in predictions for each peptide length, cleanup the netMHC output, and combine them in a single tsv file

```{r}
combine_and_clean <- function(allele){

  file.list <- list.files("./Output/netMHCpan_predictions_test/", pattern = allele)
  file.list <- paste0("./Output/netMHCpan_predictions_test/", file.list)
  
  file.list %>% 
    map(read_lines) %>% 
    map(~.[grepl("   1 HLA", .)]) %>% 
    unlist() %>% 
    c(" Pos         MHC        Peptide      Core Of Gp Gl Ip Il        Icore        Identity  Score_EL %Rank_EL", .) %>% 
    write_lines(paste0("./Output/netMHCpan_predictions_test/", allele,"_all_clean.tsv"))
  
  gc()
}
```


Apply the function for each allele

```{r}
HLA.alleles <-  c("HLA-A0201","HLA-A0101", "HLA-A2402", 
                  "HLA-B0801", "HLA-B1801","HLA-B5101", "HLA-B1501", "HLA-B4402", 
                  "HLA-C0501","HLA-C0701", "HLA-C1601", "HLA-C0304")

map(HLA.alleles, combine_and_clean)
```


## Combine files per tumor line

Write a function that reads in the allele predictions per tumor line, only imports the columns needed, joins them together, and selects the highest affinity rank.

```{r}
write_tumor_files <- function(alls, tum){
  file.list <- paste0("./Output/netMHCpan_predictions_test/", alls,"_all_clean.tsv")
  
  file.list %>% 
    map(~read_table(., col_types = cols_only(Peptide = col_character(), `%Rank_EL` = col_double()))) %>% 
    map2(alls, ~setNames(.x, c("Peptide", .y))) %>% 
    reduce(left_join, by = "Peptide") %>% 
    distinct(Peptide, .keep_all = T) %>% 
    pivot_longer(cols = !Peptide, names_to = "allele", values_to = "rank") %>% 
    group_by(Peptide) %>% 
    slice_min(rank, n=1) %>% 
    write_tsv(paste0("./Output/netMHCpan_predictions_test/HLA_affinities_",tum,".tsv"))
}
```

Define HLA typings per tumor line and run the function.

```{r}
typings <- list(M026X1 = c("HLA-A0201","HLA-A0101", "HLA-B0801", "HLA-B4402", "HLA-C0501","HLA-C0701"), 
                SKMEL95 =c("HLA-A0201","HLA-A2402", "HLA-B1801","HLA-B5101","HLA-C0501","HLA-C1601"),
                NKIRTIL006 = c("HLA-A0201", "HLA-B1501", "HLA-B4402", "HLA-C0501", "HLA-C0304" ))

typings %>% 
  map2(names(typings), ~write_tumor_files(.x, .y))

```



## Add scores to test set

```{r}
test.set <- read_tsv("./Output/test_train_sets/HLA_test_set.tsv")

test.set %>% 
  rename(Peptide = "peptide") %>% 
  nest(data = !tumor) %>% 
  mutate(peptide.affinities = map(paste0("./Output/netMHCpan_predictions_test/HLA_affinities_",tumor,".tsv"), read_tsv)) %>% 
  mutate(data2 = map2(data, peptide.affinities, ~left_join(.x, .y, by = c("Peptide")))) %>% 
  select(data2, tumor) %>% 
  unnest(data2) %>% 
  group_by(tumor) %>% 
  distinct(Peptide, .keep_all = T) -> test.set.new

write_tsv(test.set.new, "./Output/test_train_sets/HLA_test_set_complete.tsv")
```



# Train data

## Clean up files

Define function to read in predictions for each peptide length, cleanup the netMHC output, and combine them in a single tsv file

```{r}
combine_and_clean <- function(allele){

  file.list <- list.files("./Output/netMHCpan_predictions_train/", pattern = allele)
  file.list <- paste0("./Output/netMHCpan_predictions_train/", file.list)
  
  file.list %>% 
    map(read_lines) %>% 
    map(~.[grepl("   1 HLA", .)]) %>% 
    unlist() %>% 
    c(" Pos         MHC        Peptide      Core Of Gp Gl Ip Il        Icore        Identity  Score_EL %Rank_EL", .) %>% 
    write_lines(paste0("./Output/netMHCpan_predictions_train/", allele,"_all_clean.tsv"))
  
  gc()
}
```


Apply the function for each allele

```{r}
HLA.alleles <-  c("HLA-A0201","HLA-A0101", "HLA-A2402", 
                  "HLA-B0801", "HLA-B1801","HLA-B5101", "HLA-B1501", "HLA-B4402", 
                  "HLA-C0501","HLA-C0701", "HLA-C1601", "HLA-C0304")

map(HLA.alleles, combine_and_clean)
```


## Combine files per tumor line

Write a function that reads in the allele predictions per tumor line, only imports the columns needed, joins them together, and selects the highest affinity rank.

```{r}
write_tumor_files <- function(alls, tum){
  file.list <- paste0("./Output/netMHCpan_predictions_train/", alls,"_all_clean.tsv")
  
  file.list %>% 
    map(~read_table(., col_types = cols_only(Peptide = col_character(), `%Rank_EL` = col_double()))) %>% 
    map2(alls, ~setNames(.x, c("Peptide", .y))) %>% 
    reduce(left_join, by = "Peptide") %>% 
    distinct(Peptide, .keep_all = T) %>% 
    pivot_longer(cols = !Peptide, names_to = "allele", values_to = "rank") %>% 
    group_by(Peptide) %>% 
    slice_min(rank, n=1) %>% 
    write_tsv(paste0("./Output/netMHCpan_predictions_train/HLA_affinities_",tum,".tsv"))
}
```

Define HLA typings per tumor line and run the function.

```{r}
typings <- list(M026X1 = c("HLA-A0201","HLA-A0101", "HLA-B0801", "HLA-B4402", "HLA-C0501","HLA-C0701"), 
                SKMEL95 =c("HLA-A0201","HLA-A2402", "HLA-B1801","HLA-B5101","HLA-C0501","HLA-C1601"),
                NKIRTIL006 = c("HLA-A0201", "HLA-B1501", "HLA-B4402", "HLA-C0501", "HLA-C0304" ))

typings %>% 
  map2(names(typings), ~write_tumor_files(.x, .y))

```



## Add scores to test set

```{r}
train.set <- read_tsv("./Output/test_train_sets/HLA_train_set.tsv")

train.set %>% 
  rename(Peptide = "peptide") %>% 
  nest(data = !tumor) %>% 
  mutate(peptide.affinities = map(paste0("./Output/netMHCpan_predictions_train/HLA_affinities_",tumor,".tsv"), read_tsv)) %>% 
  mutate(data2 = map2(data, peptide.affinities, ~left_join(.x, .y, by = c("Peptide")))) %>% 
  select(data2, tumor) %>% 
  unnest(data2) %>% 
  group_by(tumor) %>% 
  distinct(Peptide, .keep_all = T) -> train.set.new

write_tsv(train.set.new, "./Output/test_train_sets/HLA_train_set_complete.tsv")
```


