---
title: "XGB results top 0.1% Ovarian"
author: "Kaspar Bresser"
date: "28/03/2024"

output: 
  github_document:
    toc: true
  html_document: 
    theme: simplex
    highlight: pygments
#    code_folding: show
    self_contained: TRUE
    toc: yes
    toc_float:
      collapsed: no
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message=FALSE,
                      autodep  = TRUE,
                      cache = FALSE,
#                     comment = NA,
                      fig.width = 5,
                      fig.asp = 0.618,
                      fig.align = "center")
```

Used the analysis below to compare the performance of the XGB models.

```{r loading}
library(bayestestR)
library(tidyverse)
library(rstatix)
library(ggpubr)
library(pROC)
```

## Import and tidy data

Import the results table.

```{r import_results}
XGB.results <- read_tsv("./Output/all_predictions_ovarian.tsv")
```

Convert the ligand column to a binary so it can be used for cumsum calculation, pivot to longer data by putting the models in a grouped format.

```{r prep_data}
XGB.results %>% 
  mutate(rank = -rank) %>% 
  mutate(detected = case_when(ligand == TRUE ~ 1,
                              TRUE ~ 0)) %>% 
  pivot_longer(cols = !c("detected","ligand", "tumor", "sequence", "swissprot.id"), 
               names_to =  "model", 
               values_to =  "score") -> XGB.results

XGB.results

```


For these analysis we'll focus on the top 0.1% scoring peptides for each model. Let's subset on those. Note that I arrange on both both model scores and a random number for tie breaking.

Next we'll calculate the cumulative sums by grouping by allele and model. 

Note that `cumsum()` takes an ordered dataframe. 

```{r data_prep3}
XGB.results %>% 
  group_by(tumor, model) %>% 
  mutate(random = sample(1:n())) %>% 
  arrange(desc(score), random, .by_group = T) %>% 
    mutate(peptides = 1:n()/n(), 
         detected_ligands = cumsum(detected)/sum(detected)) %>% 
  slice_head(prop = .001) -> XGB.results
 
XGB.results
```




Lastly, set the ordering in which we'd like the models to appear in plots, by converting the models variable to a factor

```{r data_prep4}
unique(XGB.results$model)

model.order <- c("aff_only", "aff_lib_small","aff_lib")

tumor.order <- as.character(unique(XGB.results$tumor))

colors <- c("#6BAED6", "#756CB1", "#5F3D8C")

XGB.results %>% 
  filter(model %in% model.order) %>% 
  mutate(model = as_factor(model),
         model = fct_relevel(model, model.order),
         tumor = as_factor(tumor),
         tumor = fct_relevel(tumor, tumor.order)) %>% 
  separate(tumor, into = c("treatment", "tumor")) -> XGB.results

XGB.results
```


## Positive predictive value

Lets first plot the positive predictive value (PPV), i.e. the number of true positives within a certain threshold. Calculate the PPV as the number of true positives divided by the total number of peptides within the selected pool.



```{r}
XGB.results %>%  
  group_by(tumor, treatment, model) %>% 
  mutate(model = fct_drop(model)) %>% 
  summarise(PPV = sum(detected)/n()) %>% 
  ungroup() %>% 
  group_by(treatment) %>% 
  rstatix::t_test(PPV ~ model, paired = T) %>% 
  add_xy_position(x = "model",  fun = "max", step.increase = 0.04) %>% 
  mutate(p.adj = round(p.adj, 3))-> stats

```



```{r PVV_plot_increment, fig.width=5, fig.asp=.85}
library(lemon)

XGB.results %>%  
  group_by(tumor, treatment, model) %>% 
  summarise(PPV = sum(detected)/n()) %>% 
   ggplot( aes(x = model, y = PPV, group = tumor)) +
    geom_boxplot(aes(group = model), width = 0.585)+
    geom_line()+
    geom_point(aes(fill = model), shape = 21, size = 3.25 )+
    facet_rep_wrap(~treatment)+
    scale_fill_manual(values = colors, labels = c("Aff", "Aff+Lib", "Aff+Lib 2.0"))+
    stat_pvalue_manual(stats, label = "p.adj", 
                     tip.length = 0.00, hide.ns = F, label.size = 2 )+
    ggtitle("PPV in top 0.1%")+
    theme_classic()+
    theme(plot.title = element_text(hjust = 0.5), 
          panel.grid.major.y = element_line(), axis.text.x=element_blank(), legend.position = "bottom",)

ggsave("Figs/XGB_PPV_strip.pdf", width = 3.2, height = 3.2, scale = 1.2)
```


## Area under the curve

Or calculate the AUC of these curves as a summary metric and plot that. Used the `area_under_curve()` function from the [bayestestR](https://easystats.github.io/bayestestR/index.html) package. 



```{r}
XGB.results %>% 
  group_by(tumor, treatment, model) %>% 
  summarise(auc = area_under_curve(peptides, detected_ligands)*1000) %>% 
  mutate(model = fct_drop(model)) %>%
  group_by(treatment) %>% 
  rstatix::t_test( auc ~ model, paired = T) %>% 
  add_xy_position(x = "model", step.increase = 0.04) %>% 
  mutate(p.adj = round(p.adj, 4)) -> stats

stats
```


```{r PVV_plot_increment2, fig.width=5, fig.asp=.85}
XGB.results %>% 
  group_by(tumor, treatment, model) %>% 
  summarise(auc = area_under_curve(peptides, detected_ligands)*1000) %>% 
    ggplot( aes(x = model, y = auc, group = tumor)) +
    geom_boxplot(aes(group = model), width = 0.585)+
    geom_line()+
    geom_point(aes(fill = model), shape = 21, size = 3.25 )+
   facet_rep_wrap(~treatment)+
    scale_fill_manual(values = colors, labels = c("Aff", "Aff+Lib", "Aff+Lib 2.0"))+
    stat_pvalue_manual(stats, label = "p.adj", 
                     tip.length = 0.00, hide.ns = F, label.size = 2 )+
    ggtitle("AUC in top 0.1%")+
    theme_classic()+
    theme(plot.title = element_text(hjust = 0.5), 
          panel.grid.major.y = element_line(), axis.text.x=element_blank(), legend.position = "bottom",)

ggsave("Figs/XGB_AUC_strip.pdf", width = 3.2, height = 3.2, scale = 1.2)
```




