---
title: "XGB results top 0.1%"
author: "Kaspar Bresser"
date: "28/02/2024"

output: 
  github_document:
    toc: true
  html_document: 
    theme: simplex
    highlight: pygments
#    code_folding: show
    self_contained: TRUE
    toc: yes
    toc_float:
      collapsed: no
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message=FALSE,
                      autodep  = TRUE,
                      cache = FALSE,
#                     comment = NA,
                      fig.width = 5,
                      fig.asp = 0.618,
                      fig.align = "center")
```

Used the analysis below to compare the performance of the XGB models.

```{r loading}
library(bayestestR)
library(tidyverse)
```

## Import and tidy data

Import the results table.

```{r import_results}
XGB.results <- read_tsv("./Output/all_predictions_new.tsv")
```

Convert the ligand column to a binary so it can be used for cumsum calculation, pivot to longer data by putting the models in a grouped format.

```{r prep_data}
XGB.results %>% 
  mutate(detected = case_when(ligand == TRUE ~ 1,
                              TRUE ~ 0)) %>% 
  mutate(rank = -rank) %>% 
  pivot_longer(cols = !c("detected", "ligand", "tumor", "swissprot_id", "Peptide", "allele"), 
               names_to =  "model", 
               values_to =  "score") -> XGB.results

XGB.results
```


For these analysis we'll focus on the top 0.1% scoring peptides for each model. Let's subset on those. Note that I arrange on both both model scores and a random number for tie breaking.



Next we'll calculate the cumulative sums by grouping by tumor and model. 

Note that `cumsum()` takes an ordered dataframe. 


```{r data_prep3}
XGB.results %>% 
  group_by(tumor, model) %>% 
  mutate(random = sample(1:n())) %>% 
  arrange(desc(score), random, .by_group = T) %>% 
    mutate(peptides = 1:n()/n(), 
         detected_ligands = cumsum(detected)/sum(detected)) %>% 
  slice_head(prop = .001) -> XGB.results

```



Lastly, set the ordering in which we'd like the models to appear in plots, by converting the models variable to a factor

```{r data_prep4}
unique(XGB.results$model)
unique(XGB.results$tumor)

model.order <- c("aff_only", "aff_lib", "aff_lib_small", "aff_lib_small_rna", "aff_rna","aff_ribo","aff_rna_ribo",
                 "aff_lib_small_ribo", "aff_lib_small_rna_ribo")


tumor.order <- c("SKMEL95", "M026X1","NKIRTIL006")

XGB.results %>% 
  mutate(model = as_factor(model),
         model = fct_relevel(model, model.order),
         tumor = factor(tumor, levels = tumor.order)) %>% 
  filter(model %in% model.order) -> XGB.results

XGB.results
```


## Positive predictive value

Lets first plot the positive predictive value (PPV), i.e. the number of true positives within a certain threshold. Calculate the PPV as the number of true positives divided by the total number of peptides within the selected pool.

Plot improvement gained with library:

```{r PVV_plot_bars_core, fig.width=6}
core.models <- c("aff_only", "aff_lib", "aff_lib_small")
#colors <- c("#27aae1", "#6675b8", "#834e9f", "#2bb673", "#2e7a3c", "cyan3", "cyan4")
colors <- c("#6BAED6", "#9E9AC8", "#756BB1")

XGB.results %>%  
  filter(model %in% core.models) %>% 
  group_by(tumor, model) %>% 
  summarise(PPV = sum(detected)/n()) %>% 
ggplot( aes(x = tumor, y = PPV, fill = model)) +
  geom_bar(stat = "identity", width=0.75, position=position_dodge(width=0.85), color = "black")+
  ggtitle("PPV in top 0.1%")+
  scale_fill_manual(values = colors, labels = c("A", "A+SF", "A+SF(trim)"))+
  ggtitle("PPV in top 0.1%")+
  theme_classic()+
  theme(plot.title = element_text(hjust = 0.5), panel.grid.major.y = element_line())

ggsave("Figs/XGB_PPV_bar_core.pdf", width = 3, height = 2, scale = 2)
```


```{r PVV_plot_bars_rna, fig.width=6}
rna.models <- c("aff_only",  "aff_lib_small", "aff_rna","aff_ribo","aff_rna_ribo")

colors <- c("#6BAED6",  "#756BB1", "#74C476", "#31A354" ,"#006D2C")

XGB.results %>%  
  filter(model %in% rna.models) %>% 
  group_by(tumor, model) %>% 
  summarise(PPV = sum(detected)/n()) %>% 
ggplot( aes(x = tumor, y = PPV, fill = model)) +
  geom_bar(stat = "identity", width=0.75, position=position_dodge(width=0.85), color = "black")+
  ggtitle("PPV in top 0.1%")+
  scale_fill_manual(values = colors, labels = c("A", "A+SF", "A+RNA", "A+RP", "A+RNA+RP"))+
  ggtitle("PPV in top 0.1%")+
  theme_classic()+
  theme(plot.title = element_text(hjust = 0.5), panel.grid.major.y = element_line())

ggsave("Figs/XGB_PPV_bar_RNA.pdf", width = 3, height = 2, scale = 2)
```



```{r PVV_plot_bars_ribo, fig.width=6}
sf.models <- c("aff_only",  "aff_lib_small", "aff_lib_small_rna","aff_lib_small_ribo","aff_lib_small_rna_ribo")

colors <- c("#6BAED6",  "#756BB1", "#2171B5", "#08519C" ,"#08306B")

XGB.results %>%  
  filter(model %in% sf.models) %>% 
  group_by(tumor, model) %>% 
  summarise(PPV = sum(detected)/n()) %>% 
ggplot( aes(x = tumor, y = PPV, fill = model)) +
  geom_bar(stat = "identity", width=0.75, position=position_dodge(width=0.85), color = "black")+
  ggtitle("PPV in top 0.1%")+
  scale_fill_manual(values = colors, labels = c("A", "A+SF", "A+SF+RNA", "A+SF+RP", "A+SF+RNA+RP"))+
  ggtitle("PPV in top 0.1%")+
  theme_classic()+
  theme(plot.title = element_text(hjust = 0.5), panel.grid.major.y = element_line())

ggsave("Figs/XGB_PPV_bar_SF.pdf", width = 3, height = 2, scale = 2)
```





```{r PVV_plot_bars_all, fig.width=6}
sf.models <- c("aff_only",  "aff_rna", "aff_lib_small_rna", "aff_ribo","aff_lib_small_ribo", "aff_rna_ribo","aff_lib_small_rna_ribo")

colors <- c("#6BAED6",  "#74C476", "#2171B5","#31A354", "#08519C", "#006D2C","#08306B")

XGB.results %>%  
  filter(model %in% sf.models) %>% 
  mutate(model = factor(model, levels = sf.models)) %>% 
  group_by(tumor, model) %>% 
  summarise(PPV = sum(detected)/n()) %>% 
ggplot( aes(x = tumor, y = PPV, fill = model)) +
  geom_bar(stat = "identity", width=0.75, position=position_dodge(width=0.85), color = "black")+
  ggtitle("PPV in top 0.1%")+
  scale_fill_manual(values = colors)+
  ggtitle("PPV in top 0.1%")+
  theme_classic()+
  theme(plot.title = element_text(hjust = 0.5), panel.grid.major.y = element_line())

ggsave("Figs/XGB_PPV_bar_SF_compared.pdf", width = 3, height = 2, scale = 2)
```




Can also plot the PPV on a continuous scale across the top 0.1%. 

```{r PPV_continuous_all, fig.width=12, fig.asp=.4}
colors <- c("#6BAED6", "#9E9AC8", "#756BB1")

XGB.results %>% 
  filter(model %in% core.models) %>% 
  group_by(tumor,model) %>% 
  mutate(PPV = cumsum(detected)/1:n()) %>% 
  ggplot(aes(x = peptides, y = PPV, color = model))+
  geom_line(linewidth = 1.2)+
  scale_color_manual(values = colors)+
  facet_wrap(~tumor, scales = "free")+
  theme_classic()+
  theme(panel.grid.major.y = element_line())

ggsave(("Figs/XGB_PPV_line_core.pdf"), width = 6, height = 2, scale = 2)
```



```{r PPV_continuous_all2, fig.width=12, fig.asp=.4}
colors <- c("#6BAED6",  "#756BB1", "#74C476", "#31A354" ,"#006D2C")

XGB.results %>% 
  filter(model %in% rna.models) %>% 
  group_by(tumor,model) %>% 
  mutate(PPV = cumsum(detected)/1:n()) %>% 
ggplot(aes(x = peptides, y = PPV, color = model))+
  geom_line(linewidth = 1.2)+
  scale_color_manual(values = colors)+
  facet_wrap(~tumor, scales = "free")+
  theme_classic()+
  theme(panel.grid.major.y = element_line())

ggsave(("Figs/XGB_PPV_line_rna.pdf"), width = 6, height = 2, scale = 2)
```


```{r PPV_continuous_all3, fig.width=12, fig.asp=.4}
sf.models <- c("aff_only",  "aff_lib_small", "aff_lib_small_rna","aff_lib_small_ribo","aff_lib_small_rna_ribo")

colors <- c("#6BAED6",  "#756BB1", "#2171B5", "#08519C" ,"#08306B")

XGB.results %>% 
  filter(model %in% sf.models) %>% 
    group_by(tumor,model) %>% 
  mutate(PPV = cumsum(detected)/1:n()) %>% 
ggplot(aes(x = peptides, y = PPV, color = model))+
  geom_line(linewidth = 1.2)+
  scale_color_manual(values = colors)+
  facet_wrap(~tumor, scales = "free")+
  theme_classic()+
  theme(panel.grid.major.y = element_line())

ggsave(("Figs/XGB_PPV_line_SF.pdf"), width = 6, height = 2, scale = 2)
```

## Cumulative sums



Core models:

```{r plot_cumsum_core, fig.width=12, fig.asp=.4}
colors <- c("#6BAED6", "#9E9AC8", "#756BB1")


XGB.results %>% 
  filter(model %in% core.models) %>% 
  ggplot(aes(x = peptides, y = detected_ligands, color = model))+
  geom_line(size = 1.2)+  
  scale_color_manual(values = colors)+
  facet_wrap(~tumor, scales = "free")+
  theme_classic()+
  theme(panel.grid.major.y = element_line())

ggsave(("Figs/XGB_cumsum_core.pdf"), width = 6, height = 2, scale = 2)
```

```{r plot_cumsum_core2, fig.width=12, fig.asp=.4}
colors <- c("#6BAED6",  "#756BB1", "#74C476", "#31A354" ,"#006D2C")


XGB.results %>% 
  filter(model %in% rna.models) %>% 
  ggplot(aes(x = peptides, y = detected_ligands, color = model))+
  geom_line(size = 1.2)+  
  scale_color_manual(values = colors)+
  facet_wrap(~tumor, scales = "free")+
  theme_classic()+
  theme(panel.grid.major.y = element_line())

ggsave(("Figs/XGB_cumsum_rna.pdf"), width = 6, height = 2, scale = 2)
```


```{r plot_cumsum_core3, fig.width=12, fig.asp=.4}
colors <- c("#6BAED6", "#9E9AC8", "#756BB1")
colors <- c("#6BAED6",  "#756BB1", "#74C476", "#31A354" ,"#006D2C")
colors <- c("#6BAED6",  "#756BB1", "#2171B5", "#08519C" ,"#08306B")


XGB.results %>% 
  filter(model %in% sf.models) %>% 
  ggplot(aes(x = peptides, y = detected_ligands, color = model))+
  geom_line(size = 1.2)+  
  scale_color_manual(values = colors)+
  facet_wrap(~tumor, scales = "free")+
  theme_classic()+
  theme(panel.grid.major.y = element_line())

ggsave(("Figs/XGB_cumsum_SF.pdf"), width = 6, height = 2, scale = 2)
```


## Area under the curve

Or calculate the AUC of these curves as a summary metric and plot that. Used the `area_under_curve()` function from the [bayestestR](https://easystats.github.io/bayestestR/index.html) package. 
First check maximum AUC to normalize between 0 and 1.



Core models:

```{r plot_auc_core1, fig.width=10, fig.asp=.45}
colors <- c("#6BAED6", "#9E9AC8", "#756BB1")


XGB.results %>% 
  filter(model %in% core.models) %>% 
  group_by(tumor, model) %>% 
  summarise(auc = area_under_curve(peptides, detected_ligands)) %>% 
  ggplot( aes(x = tumor, y = auc, fill = model)) +
  geom_bar(stat = "identity", width=0.75, position=position_dodge(width=0.85), color = "black")+  
  scale_fill_manual(values = colors, labels = c("Aff", "Aff+Lib", "Aff+Lib(trim)", "Aff+RNA"))+
  ggtitle("AUC in top 0.1%")+
  theme_classic()+
  theme(plot.title = element_text(hjust = 0.5), panel.grid.major.y = element_line())

ggsave(("Figs/XGB_AUC_core.pdf"), width = 3, height = 2, scale = 2)
```


```{r plot_auc_core2, fig.width=10, fig.asp=.45}
colors <- c("#6BAED6",  "#756BB1", "#74C476", "#31A354" ,"#006D2C")

XGB.results %>% 
  filter(model %in% rna.models) %>% 
  group_by(tumor, model) %>% 
  summarise(auc = area_under_curve(peptides, detected_ligands)) %>% 
  ggplot( aes(x = tumor, y = auc, fill = model)) +
  geom_bar(stat = "identity", width=0.75, position=position_dodge(width=0.85), color = "black")+  
  scale_fill_manual(values = colors)+
  ggtitle("AUC in top 0.1%")+
  theme_classic()+
  theme(plot.title = element_text(hjust = 0.5), panel.grid.major.y = element_line())

ggsave(("Figs/XGB_AUC_rna.pdf"), width = 3, height = 2, scale = 2)
```



```{r plot_auc_core3, fig.width=10, fig.asp=.45}
colors <- c("#6BAED6",  "#756BB1", "#2171B5", "#08519C" ,"#08306B")

XGB.results %>% 
  filter(model %in% sf.models) %>% 
  group_by(tumor, model) %>% 
  summarise(auc = area_under_curve(peptides, detected_ligands)) %>% 
  ggplot( aes(x = tumor, y = auc, fill = model)) +
  geom_bar(stat = "identity", width=0.75, position=position_dodge(width=0.85), color = "black")+  
  scale_fill_manual(values = colors)+
  ggtitle("AUC in top 0.1%")+
  theme_classic()+
  theme(plot.title = element_text(hjust = 0.5), panel.grid.major.y = element_line())

ggsave(("Figs/XGB_AUC_sf.pdf"), width = 3, height = 2, scale = 2)
```

```{r plot_auc_core4, fig.width=10, fig.asp=.45}
sf.models <- c("aff_only",  "aff_rna", "aff_lib_small_rna", "aff_ribo","aff_lib_small_ribo", "aff_rna_ribo","aff_lib_small_rna_ribo")

colors <- c("#6BAED6",  "#74C476", "#2171B5","#31A354", "#08519C", "#006D2C","#08306B")

XGB.results %>% 
  filter(model %in% sf.models) %>% 
    mutate(model = factor(model, levels = sf.models)) %>% 
  group_by(tumor, model) %>% 
  summarise(auc = area_under_curve(peptides, detected_ligands)*1000) %>% 
  ggplot( aes(x = tumor, y = auc, fill = model)) +
  geom_bar(stat = "identity", width=0.75, position=position_dodge(width=0.85), color = "black")+  
  scale_fill_manual(values = colors)+
  ggtitle("AUC in top 0.1%")+
  theme_classic()+
  theme(plot.title = element_text(hjust = 0.5), panel.grid.major.y = element_line())

ggsave(("Figs/XGB_AUC_sf_compared.pdf"), width = 3, height = 2, scale = 2)
```



## Sensitivity ect 

Read in the data to get metrics of model performance

```{r}
library(pROC)

XGB.results <- read_tsv("./Output/all_predictions_new.tsv")

XGB.results <- mutate(XGB.results, rank = -rank)
```


Define a function to retrieve the performance metrics, and a function to get the cutoffs at a certain threshold of sensitivity

```{r}
get_sens_spec <- function(lig, preds, mods){
 
  map(preds, ~roc(lig, .)) %>% 
     map(~coords(., x = "all", input = "sensitivity", ret = "all") ) -> tmp

    
  tibble(model = mods, dat = tmp) %>% 
    unnest()
    
}

get_cutoffs <- function(lig, preds, mods, thresh){
  map(preds, ~roc(lig, .)) %>%
    map(~coords(., x = thresh, input = "sensitivity", ret = "all") ) %>% 
  map(~select(., !c(threshold))) %>% 
  bind_rows() %>% 
    mutate(model = mods)
}
```

Get the information

```{r}
model.results.info <- get_sens_spec(lig = XGB.results$ligand,
                                    preds = map2(list(XGB.results), model.order, ~pull(.x, .y)),
                                    mods =  model.order)


coordinates <- get_cutoffs(lig = XGB.results$ligand,
                           preds = map2(list(XGB.results), model.order, ~pull(.x, .y)),
                           mods =  model.order,
                           thresh = 0.1)

model.results.info
```

And plot at 0.1 sensitivity

```{r}
colors <- c("#6BAED6",  "#756BB1", "#74C476", "#31A354" ,"#006D2C")

model.results.info %>% 
  filter(model %in% rna.models) %>% 
  mutate(model = factor(model, levels = rna.models)) %>% 
ggplot( aes(x = `1-specificity`, y = sensitivity, color = model))+
  geom_line( linewidth = 1.5)+
  coord_cartesian(ylim = c(0.0, 0.1), xlim = c(0.0, 0.001))+
  geom_segment(aes(x = 0.0, xend = max(coordinates$`1-specificity`), y = .1, yend = .1), color = "black", linetype = "dashed")+
  geom_segment(data = filter(coordinates, model %in% rna.models), aes(x = `1-specificity`, xend = `1-specificity`, y = 0.0, yend = .1, color = model), 
               linewidth = 0.8, linetype = "dashed")+
  scale_color_manual(values = colors)+
  theme_classic()+
  theme(panel.grid.major = element_line())

ggsave(("Figs/XGB_SensSpec_10.pdf"), width = 2.8, height = 2, scale = 2)

```


And plot at 0.5 sensitivity


```{r}
coordinates <- get_cutoffs(lig = XGB.results$ligand,
                           preds = map2(list(XGB.results), model.order, ~pull(.x, .y)),
                           mods =  model.order,
                           thresh = 0.5)

colors <- c("#6BAED6",  "#756BB1", "#74C476", "#31A354" ,"#006D2C")

model.results.info %>% 
  filter(model %in% rna.models) %>% 
  mutate(model = factor(model, levels = rna.models)) %>% 
ggplot( aes(x = `1-specificity`, y = sensitivity, color = model))+
  geom_line( linewidth = 1.5)+
  coord_cartesian(ylim = c(0.0, 0.5), xlim = c(0.0, 0.015))+
  geom_segment(aes(x = 0.0, xend = max(coordinates$`1-specificity`), y = .5, yend = .5), color = "black", linetype = "dashed")+
  geom_segment(data = filter(coordinates, model %in% rna.models), aes(x = `1-specificity`, xend = `1-specificity`, y = 0.0, yend = .5, color = model), 
               linewidth = 0.8, linetype = "dashed")+
  scale_color_manual(values = colors)+
  theme_classic()+
  theme(panel.grid.major = element_line())

ggsave(("Figs/XGB_SensSpec_50.pdf"), width = 2.8, height = 2, scale = 2)
```

